---
title: "Assignment 3"
author: "Evan Hollier"
date: "2023-03-01"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

```{r include=FALSE }
knitr::opts_chunk$set(echo = TRUE)
#install.packages("RMySQL")
library(RMySQL)
library(ggplot2)
library(HistData)
```

Connect to the mlb database.

```{r}
db = dbConnect(MySQL(), 
                 user='root', 
                 password='mysql', 
                 dbname='mlb', 
                 host='localhost')

dbListTables(db)
```

**1) Plot a histogram of the players age as of January 1st, 2014. When the question asks for players, be sure and exclude the managers.**

```{r}
rs = dbSendQuery(db, 'SELECT player_id, birth_year, birth_month, birth_day 
                      FROM mlb_master 
                      WHERE player_id NOT IN
                      (SELECT player_id FROM mlb_manager)'
                 )
dat = dbFetch(rs, n = -1) # n = -1 fetches all rows
dat$age = as.integer(difftime(
  "2014-1-1",
  paste(dat$birth_year,dat$birth_month,dat$birth_day, sep='-')))
dat$age = dat$age / 365
ggplot(dat, aes(x=age)) +
        geom_histogram(binwidth=1) +
        ggtitle("Player Age")
```

**2) Do a regression on players heights and weights to get the mathematical relationship between the two. (get the slope and intercept of the regression).**\
(Done in Python)\
Y = -154.89 + 4.9455X\
\
**3) Plot the relationship between players age (as of Jan. 1, 2014) and batting average. Batting average is determined by dividing a player's hits by his total at-bats for a number between zero (shown as .000) and one (1.000). Do players get better or worse with age? What is the best age for a player?**

```{r}
rs = dbSendQuery(db, 'SELECT player_id, hits/at_bats AS batting_avg, team_id,
                 hits, at_bats  
                 FROM mlb_batting 
                 WHERE at_bats>1 AND
                 player_id NOT IN
                 (SELECT player_id FROM mlb_manager)'
                 )
# Players whose have never been at bat are filtered.
dat1 = dbFetch(rs, n = -1)
dat1 <- merge(dat[c("player_id","age")], dat1, by="player_id")

ggplot(dat1, aes(x = age, y = batting_avg)) +
  geom_point() +
  stat_smooth(method='lm')

# Determine best age:
dat1$flat_age = floor(dat1$age)
batavg_age <- aggregate(hits~flat_age, dat1, sum)
batavg_age <- merge(aggregate(at_bats~flat_age, dat1, sum), 
                    batavg_age, by="flat_age")
batavg_age$batting_avg = batavg_age$hits / batavg_age$at_bats

batavg_age[batavg_age$batting_avg == max(batavg_age$batting_avg),]
```

Using linear regression, there doesn't seem to be any relationship between age and batting average. The line of regression is horizontal, meaning as age increases, batting average doesn't change. The age with the best batting average is 20.

**4) Do heavier players hit more home runs per at bat than lighter players? Plot the relationship between weight vs home runs per at bat. Compare home runs per at bat for above average weight players to below average weight players? Is the difference statistically significant? Remember when calculating the average, don't average the averages. Sum up the home runs, sum up the at bats, then divide them.**

```{r}
rs = dbSendQuery(db, 'SELECT weight, 
                 homeruns, at_bats
                 FROM mlb_master
                 INNER JOIN mlb_batting
                 ON mlb_master.player_id = mlb_batting.player_id
                 WHERE at_bats>20 AND
                 mlb_master.player_id NOT IN
                 (SELECT player_id FROM mlb_manager)'
                 )
dat2 = dbFetch(rs, n = -1)

ggplot(dat2, aes(x = weight, y = homeruns/at_bats)) +
  geom_jitter(size=0.6) +
  stat_smooth(method='lm')

# Compare above average to below average:
avg_weight <- mean(dat2$weight)
homerunavg_weight <- aggregate(homeruns~weight>avg_weight, dat2, sum)
homerunavg_weight <- merge(aggregate(at_bats~weight>avg_weight, dat2, sum),   
                           homerunavg_weight)
homerunavg_weight$homerun_avg = homerunavg_weight$homeruns / homerunavg_weight$at_bats

# Determine statistical significance
true_homerun_avg = sum(homerunavg_weight$homeruns) / sum(homerunavg_weight$at_bats)
dat2.below <- dat2[dat2$weight < avg_weight,]
dat2.below$homerun_avg = dat2.below$homeruns / dat2.below$at_bats
dat2.above <- dat2[dat2$weight > avg_weight,]
dat2.above$homerun_avg = dat2.above$homeruns / dat2.above$at_bats

t.test(dat2.below$homerun_avg, 
       dat2.above$homerun_avg,
       mu=true_homerun_avg)
```

Using linear regression, the relationship between weight and home run average is shown to be positive. The line of regression slopes upward, meaning as weight increases, home run average increases. The home run average is 0.01906026 for below average weight players, and 0.03176666 for above average. The home run average for all players is 0.02550918. Performing a Two Sample t-test on the home run averages above and below average weight indicates the difference in means is statistically significant.

**5) Which team has the oldest average age for players? Is there a relationship between average age of players and winning percentage?**

```{r}
rs = dbSendQuery(db, 'SELECT player_id, mlb_batting.team_id FROM mlb_batting
                      INNER JOIN mlb_team
                      ON mlb_team.team_id = mlb_batting.team_id
                      WHERE mlb_batting.player_id NOT IN
                      (SELECT player_id FROM mlb_manager)'
                 )
dat3 = dbFetch(rs, n = -1)
dat3 <- merge(dat[c("player_id","age")], dat3, by="player_id")
dat3 <- aggregate(age~team_id, dat3, mean)
dat3[dat3$age == max(dat3$age),]

rs = dbSendQuery(db, 'SELECT team_id, wins/games as winrate FROM mlb_team')
dat3.2 = dbFetch(rs, n = -1)
dat3 <- merge(dat3, dat3.2, by="team_id")

ggplot(dat3, aes(x = age, y = winrate)) +
  geom_point() +
  stat_smooth(method='lm')
```

New York Yankees (NYA) have the oldest average age for players. Using linear regression, the relationship between a team's average age and winning percentage is shown to be positive. The line of regression slopes upward, meaning as average age increases, winning percentage increases.

**6) Calculate the percentage of times a player is caught stealing (from batting table: caught_stealing/(stolen_bases + caught_stealing) \* 100.\
Is there a statistically significant difference between the best manager and the worst manager? Best manager being defined as the manager with the lowest percentage of times their players get caught stealing.**

```{r}
rs = dbSendQuery(db, 'SELECT team_id, stolen_bases, caught_stealing
                      FROM mlb_batting
                      WHERE mlb_batting.player_id NOT IN
                      (SELECT player_id FROM mlb_manager)'
                 )
dat4 = dbFetch(rs, n = -1)
dat4.ag <- aggregate(stolen_bases~team_id, dat4, sum)
dat4.ag <- merge(aggregate(caught_stealing~team_id, dat4, sum),   
                           dat4.ag)
# Impossible to differentiate player stats based on manager stint, 
# so stick with just team_id to represent a manager.
dat4.ag$caught_rate = dat4.ag$caught_stealing / (dat4.ag$stolen_bases + dat4.ag$caught_stealing) * 100
dat4.ag

min(dat4.ag$caught_rate)
max(dat4.ag$caught_rate)
t.test(dat4.ag$caught_rate)
```

There is a statistically significant difference between the best manager and the worst manager. The best and worst managers' percentages of times their players get caught stealing are both well outside the 95 percent confidence interval for all managers.\
\
**7) For each team, calculate the batting average (hits/at_bats) for the team. Do all teams with better than average batting also have better than 50% winning percentage? Print a table with all the teams that have above average batting and less than 50% winning percentage.**

```{r}
dat5 <- aggregate(hits~team_id, dat1, sum)
dat5 <- merge(aggregate(at_bats~team_id, dat1, sum), dat5, by="team_id")
dat5$batting_avg = dat5$hits / dat5$at_bat
dat5 <- merge(dat5[c("team_id", "batting_avg")], 
              dat3[c("team_id", "winrate")], 
              by="team_id")

(true_batting_avg <- mean(dat5$batting_avg))
dat5 <- dat5[dat5$batting_avg > true_batting_avg,]
dat5 <- dat5[dat5$winrate < .5,]
dat5
```

No, there are some teams with above average batting and less than 50% winning percentage. These teams are: CHA, COL, MIA, MIN, and TEX.
